\section{ICY metadata}
\emph{ICY metadata} is the name for the mechanism used to update
metadata in Icecast's source streams.
The techniques is primaly intented for data formats that do not support in-stream
metadata, such as mp3 or AAC. However, it appears that Icecast also supports
ICY metadata update for ogg/vorbis streams.

When using the ICY metadata update mechanism, new metadata are submited seperatly from
the stream's data, via a http GET request. The format of the request depends on the
protocol you are using (ICY for shoutcast and icecast 1 or HTTP for icecast 2).

Starting with 1.0, you can do several interesting things with icy metadata updates
in liquidsoap. We list some of those here.

\subsection{Enable/disable ICY metadata updates}
You can enable or disable icy metadata update in \verb+output.icecast+
by setting the \verb+icy_metadata+ parameter to either \verb+"true"+
or \verb+"false"+. The default value is \verb+"auto"+ and does the following:

\begin{itemize}
\item Set \verb+"true"+ for: mp3, aac, aac+, wav
\item Set \verb+"false"+ for any format using the ogg container

\end{itemize}
You may, for instance, enable icy metadata update for ogg/vorbis
streams.

\subsection{Update metadata manually}
The function \verb+icy.update_metadata+ implements a manual metadata update
using the ICY mechanism. It can be used independently from the \verb+icy_metadata+
parameter described above, provided icecast supports ICY metadata for the intented stream.

For instance the following script registers a telnet command name \verb+metadata.update+ 
that can be used to manually update metadata:

\begin{verbatim}
def icy_update(v) =
  # Parse the argument
  l = string.split(separator=",",v)
  def split(l,v) =
    v = string.split(separator="=",v)
    if list.length(v) >= 2 then
      list.append(l,[(list.nth(v,0),list.nth(v,1))])
    else
      l
    end
  end
  meta = list.fold(split,[],l)

  # Update metadata
  icy.update_metadata(mount="/mystream",password="hackme",
                      host="myserver.net",meta)
  "Done !"
end

server.register("update",namespace="metadata",
                 description="Update metadata",
                 usage="update title=foo,album=bar,..",
                 icy_update)
\end{verbatim}
As ususal, \verb+liquidsoap -h icy.update_metadata+ lists all the arguments
of the function.

